Reasonably Programmable Literal Notation

### 2018-06-12 icfp18/syntax-icfp18.tex
ACM copyright stuff, CCS stuff, made section titles all caps (blagh)

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

@@

\vspace{-5px}

@@

Compared to existing approaches, TLMs are uniquely \emph{reasonable}. TLM clients can reason abstractly, i.e. without examining grammars or generated expansions, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal body, which gives the locations and types of spliced subterms. TLM providers can reason modularly about syntactic ambiguity and expansion correctness according to clear criteria. This paper incorporates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system. We discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness. Finally, we establish these abstract reasoning principles formally with a detailed type-theoretic account of expression and pattern TLMs for ``core ML''. 

@@

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10011007.10011006.10011008</concept_id>
% <concept_desc>Software and its engineering~General programming languages</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003521.10003525</concept_id>
% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~General programming languages}
% \ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

@@

\vspace{-5px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
\vspace{-2px}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define shorthand \emph{literal notation}, i.e. notation that decreases the syntactic cost of constructing and pattern matching over values of some particular data structure or parameterized family of data structures. For example, many languages in the ML family support list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Lists are common across problem domains, but other literal notation is more specialized. For example, Ur/Web extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}. The example in Fig.~\ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

\begin{figure}[h]
\vspace{-9px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-4px}
\caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-7px}
\end{figure}

@@

\newcommand{\seTLMsSec}{Expression TLMs in Reason}
\section{\protect\seTLMsSec}
\label{sec:setlms}

@@

\newcommand{\spTLMsSec}{Pattern TLMs}
\section{\protect\spTLMsSec}
\label{sec:sptsms}

@@

\newcommand{\pTLMsSec}{Parametric TLMs}
\section{\protect\pTLMsSec}
\label{sec:ptsms}

@@

\newcommand{\seTLMsFormallySec}{Simple TLMs, Formally}
\section{\protect\seTLMsFormallySec}
\label{sec:setlms-formally}

@@

% % \begingroup
% % \def\thetheorem{\ref{thm:tsc-SES}}
% % \begin{theorem}[seTLM Abstract Reasoning Principles]
% % If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
% % \begin{enumerate}
% % \item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
% % \item $\encodeBody{b}{\ebody}$
% % \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
% % \item $\decodeCondE{\ecand}{\ce}$
% % \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
% % \item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
% % \item \textbf{(Typing 2)} $\sseq{
% %       \expandsTU{\uDD{\uD}{\Delta}}
% %       {
% %         \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
% %       }{\tau'_i}
% %     }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
% % \item \textbf{(Typing 3)} $\sseq{
% %   \cvalidT{\emptyset}{
% %     \tsceneUP
% %       {\uDD
% %         {\uD}{\Delta}
% %       }{b}
% %   }{
% %     \ctau_i
% %   }{\tau_i}
% % }{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
% % \item \textbf{(Typing 4)} $\sseq{
% %   \expandsU
% %     {\uDD{\uD}{\Delta}}
% %     {\uGG{\uG}{\Gamma}}
% %     {\uPsi}
% %     {\parseUExpF{\bsubseq{b}{m_i}{n_i}}}
% %     {e_i}
% %     {\tau_i}
% % }{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
% % \item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
% % \item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
% %   % $\hastypeU
% %   % {\sseq{\Dhyp{t_i}}{\nty}}
% %   % {\sseq{x_i : \tau_i}{\nexp}}
% %   % {e'}{\tau}$
% % \end{enumerate}
% % \end{theorem}
% % \begin{proof} The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in Appendix \ref{appendix:SES-reasoning-principles}.
% % \end{proof}
% % \endgroup
\newcommand{\implSec}{Implementation}
\section{\protect\implSec}
\label{sec:implementation}
The implementation, \li{Relit}, which is an ongoing effort based on the finished design described in this paper, consists of a few changes to the context-free grammar of Reason together with a parse tree pre-processor for the OCaml compiler, currently using its PPX system (discussed in Sec.~\ref{sec:existing-approaches}).

@@

\newcommand{\rwSec}{Related Work}
\section{\protect\rwSec}
\label{sec:existing-approaches}

@@

\newcommand{\discussionSec}{Discussion}
\section{\protect\discussionSec}
\label{sec:discussion}
\label{sec:conclusion}

>>>>

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
% \setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

@@

@@

Compared to existing approaches, TLMs are uniquely \emph{reasonable}. TLM clients can reason abstractly, i.e. without examining grammars or generated expansions, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal body, which gives the locations and types of spliced subterms. TLM providers can reason modularly about syntactic ambiguity and expansion correctness according to clear criteria. This paper incorporates TLMs into Reason, an emerging alternative front-end for OCaml, and demonstrates, through several non-trivial case studies, how TLMs integrate with the advanced features of OCaml, including pattern matching and the module system. We also discuss optional integration with MetaOCaml, which allows TLM providers to be more confident about type correctness. Finally, we establish these abstract reasoning principles formally with a detailed type-theoretic account of expression and pattern TLMs for ``core ML''. 

@@

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011019</concept_id>
<concept_desc>Software and its engineering~Extensible languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011050.10011055</concept_id>
<concept_desc>Software and its engineering~Macro languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Extensible languages}
\ccsdesc[500]{Software and its engineering~Macro languages}

@@

\section{Introduction}
\label{sec:intro}
When designing the surface syntax of a general-purpose programming language, it is common practice to define shorthand \emph{literal notation}, i.e. notation that decreases the syntactic cost of constructing and pattern matching over values of some particular data structure or parameterized family of data structures. For example, many languages in the ML family support list literals like \li{[x1, x2, x3]} in both expression and pattern position \cite{harper1997programming,mthm97-for-dart}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Lists are common across problem domains, but other literal notation is more specialized. For instance, Ur/Web extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data~\cite{conf/popl/Chlipala15}. For example, Fig.~\ref{fig:urweb} shows two Ur/Web HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} and the other an HTML expression delimited by \li{\{} and \li{\}}.
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

\begin{figure}[h]
\vspace{-9px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\caption{HTML literals with support for splicing at two types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-7px}
\end{figure}

@@

\section{Expression Literals}
\label{sec:setlms}

@@

\section{Pattern Literals}
\label{sec:sptsms}

@@

\section{Parametric TLMs}
\label{sec:ptsms}

@@

\section{Typed Literal Macros, Formally}
\label{sec:setlms-formally}

@@

% % \begingroup
% % \def\thetheorem{\ref{thm:tsc-SES}}
% % \begin{theorem}[seTLM Abstract Reasoning Principles]
% % If $\expandsU{\uDD{\uD}{\Delta}}{\uGG{\uG}{\Gamma}}{\uPsi}{\utsmap{\tsmv}{b}}{e}{\tau}$ then:
% % \begin{enumerate}
% % \item (\textbf{Typing 1}) $\uPsi = \uPsi', \uShyp{\tsmv}{a}{\tau}{\eparse}$ and $\hastypeU{\Delta}{\Gamma}{e}{\tau}$
% % \item $\encodeBody{b}{\ebody}$
% % \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\lbltxt{SuccessE}}{\ecand}}$
% % \item $\decodeCondE{\ecand}{\ce}$
% % \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
% % \item $\segof{\ce} = \sseq{\acesplicedt{m'_i}{n'_i}}{\nty} \cup \sseq{\acesplicede{m_i}{n_i}{\ctau_i}}{\nexp}$
% % \item \textbf{(Typing 2)} $\sseq{
% %       \expandsTU{\uDD{\uD}{\Delta}}
% %       {
% %         \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
% %       }{\tau'_i}
% %     }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
% % \item \textbf{(Typing 3)} $\sseq{
% %   \cvalidT{\emptyset}{
% %     \tsceneUP
% %       {\uDD
% %         {\uD}{\Delta}
% %       }{b}
% %   }{
% %     \ctau_i
% %   }{\tau_i}
% % }{\nexp}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\nexp}$
% % \item \textbf{(Typing 4)} $\sseq{
% %   \expandsU
% %     {\uDD{\uD}{\Delta}}
% %     {\uGG{\uG}{\Gamma}}
% %     {\uPsi}
% %     {\parseUExpF{\bsubseq{b}{m_i}{n_i}}}
% %     {e_i}
% %     {\tau_i}
% % }{\nexp}$ and $\sseq{\hastypeU{\Delta}{\Gamma}{e_i}{\tau_i}}{\nexp}$
% % \item (\textbf{Capture Avoidance}) $e = [\sseq{\tau'_i/t_i}{\nty}, \sseq{e_i/x_i}{\nexp}]e'$ for some $\sseq{t_i}{\nty}$ and $\sseq{x_i}{\nexp}$ and $e'$
% % \item (\textbf{Context Independence}) $\mathsf{fv}(e') \subset \sseq{t_i}{\nty} \cup \sseq{x_i}{\nexp}$
% %   % $\hastypeU
% %   % {\sseq{\Dhyp{t_i}}{\nty}}
% %   % {\sseq{x_i : \tau_i}{\nexp}}
% %   % {e'}{\tau}$
% % \end{enumerate}
% % \end{theorem}
% % \begin{proof} The proof, which involves auxiliary lemmas about the decomposition of proto-types and proto-expressions, is given in Appendix \ref{appendix:SES-reasoning-principles}.
% % \end{proof}
% % \endgroup
\section{Implementation}
\label{sec:implementation}
The implementation, \li{Relit}, which is an ongoing effort based on the finished design described in this paper, consists of a few changes to the context-free grammar of Reason together with a parse tree pre-processor for the OCaml compiler, currently using its PPX system (discussed in Sec.~\ref{sec:existing-approaches}).

@@

\section{Related Work}
\label{sec:existing-approaches}

@@

\section{Discussion}
\label{sec:discussion}
\label{sec:conclusion}
